---
title: "activity-based-courses  .csv"
output: 
  rmdformats::html_docco:
    css: 'codebooks.css'
bibliography: '../../notes/epsy8251.bib'
csl: '../../notes/apa-single-spaced.csl'
---


```{r knitr_init, echo=FALSE}
library(knitr)
library(rmdformats)
library(fontawesome)

## Global options
options(max.print="75")
opts_chunk$set(echo=FALSE,
	             cache=TRUE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=75)
```


The data in [activity-based-courses.csv](https://github.com/zief0002/epsy-8251/raw/master/data/activity-based-courses.csv) are synthetic data made available by Robert delMas and contain data collected from an experimental study designed to study the effects of activity-based courses on student learning. Students were randomly assigned to enroll in either the activity-based (treatment) or the lecture-based (control) section of a course. The three attributes collectd from the $n=110$ students include:

- `grade_points`: Course GPA based on the student's final grade
- `condition`: Experimental condition indicating whether the student was randomly assigned to the activity-based section (`treatment`) or the lecture-based section (`control`)
- `act`: Student's composite ACT score. Prior research has established that final course grades are correlated with ACT scores


### Preview

```{r}
activity = readr::read_csv(file = "~/Documents/github/epsy-8251/data/activity-based-courses.csv")
head(activity)
```


### References
