---
title: "colleges-bordering-mn.csv"
output: 
  rmdformats::html_docco:
    css: 'codebooks.css'
bibliography: '../../notes/epsy8251.bib'
csl: '../../notes/apa-single-spaced.csl'
---


```{r knitr_init, echo=FALSE, cache=FALSE}
library(knitr)
library(rmdformats)
library(fontawesome)

## Global options
options(max.print="75")
opts_chunk$set(echo=FALSE,
	             cache=TRUE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=75)
```


The data in [colleges-bordering-mn.csv](https://github.com/zief0002/epsy-8251/raw/master/data/colleges-bordering-mn.csv) contains institutional data for 104 colleges and universities in the five state area (MN, IA, WI, ND, and SD). These data were collected by the Department of Education for the [2013 College Score Card](https://collegescorecard.ed.gov/). The variables are:

- `name`: Name of college/university
- `tuition_in_state`: In-state tuition and fees
- `tuition_out_state`: Out-of-state tuition and fees
- `state`: State postal abbreviation (IA = Iowa, MN = Minnesota, ND = North Dakota, SD = South Dakota, WI = Wisconsin)
- `public`: Dummy-coded variable indicating educational sector (1 = public school, 0 = private school)
- `main`: Dummy-coded variable indicating whether the campus is the school's main campus? (1 = yes, 0 = no)
- `admission`: Admission rate
- `act75`: 75th percentile of the ACT cumulative scores
- `avg_fac_salary`: Average faculty salary (per month)
- `completion`: Four-year completion rate for first-time, full-time students
- `pct_pell`: Percentage of undergraduates who receive a Pell grant


### Preview

```{r}
colleges = readr::read_csv(file = "~/Documents/github/epsy-8251/data/colleges-bordering-mn.csv")
head(colleges)
```


### References
