---
title: "state-education.csv"
output: 
  rmdformats::html_docco:
    css: 'codebooks.css'
    # includes:
    #   after_body: doc-suffix.html
bibliography: '../../notes/epsy8251.bib'
csl: '../../notes/apa-single-spaced.csl'
---


```{r knitr_init, echo=FALSE, cache=FALSE}
library(knitr)
library(rmdformats)
library(fontawesome)

## Global options
options(max.print="75")
opts_chunk$set(echo=FALSE,
	             cache=FALSE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=75)
```


This data, stored in [state-education.csv](https://github.com/zief0002/epsy-8251/raw/master/data/state-education.csv) includes state-level aggregate data on six attributes. The attributes, collected for all 50 states and the District of Columbia, are:

- `state`: State name
- `postal`: State postal code
- `region`: Region of the country identified by the National Education Association (Far West, Great Lakes, Mid East, New England, Plains, Rocky Mountains, Southeast, Southwest)
- `sat_participate`: Percentage of students in the state who took the SAT.
- `sat_ebrw`: Average score on the Evidence-Based Reading and Writing (EBRW) scale in the state.
- `sat_math`: Average score on the math scale in the state.
- `sat_total`: Average total SAT score in the state.
- `salary_2020_21`: Average 2020--2021 public teacher salary in the state.


Note: All of the SAT data is based on students in the class of 2020 who took the current SAT during high school. The SAT is made up of three sections: (1) Reading, (2) Writing and Language (also just called Writing), and (3) Math. The Math section is scored on a scale of 200--800. The Reading and Writing sections are combined to give a Evidence-Based Reading and Writing (EBRW) score, also ranging from 200--800. By combining the Math and EBRW scores, we get a total SAT score ranging from 400--1600.




### Preview

```{r}
options(dplyr.width = Inf, width = 85)
state_educ = readr::read_csv(file = "~/Documents/github/epsy-8251/data/state-education.csv")
print(state_educ, width = Inf)
```


### References

College Board. (2020). [SAT suite of assessments annual report](https://reports.collegeboard.org/pdf/2020-total-group-sat-suite-assessments-annual-report.pdf). Author.

NEA Research. (2021). [Rankings of the states 2020 and estimates of school statistics 2021](https://www.nea.org/sites/default/files/2021-04/2021%20Rankings_and_Estimates_Report.pdf). National Education Association.

