---
title: "waffle-house.csv"
output: 
  rmdformats::html_docco:
    css: 'codebooks.css'
    # includes:
    #   after_body: doc-suffix.html
bibliography: '../../notes/epsy8251.bib'
csl: '../../notes/apa-single-spaced.csl'
---


```{r knitr_init, echo=FALSE, cache=FALSE}
library(knitr)
library(rmdformats)
library(fontawesome)

## Global options
options(max.print="75")
opts_chunk$set(echo=FALSE,
	             cache=TRUE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=75)
```


This data, stored in [waffle-house.csv](https://github.com/zief0002/epsy-8251/raw/master/data/waffle-house.csv) includes state-level data on 11 attributes. The attributes, collected for all 50 states are:

- `state`: State name
- `postal`: State postal code
- `divorce`: 2009 divorce rate (per 1000 adults)
- `divorce_se`: Standard error of divorce rate
- `population`: 2010 population (in millions)
- `median_age_marriage`: 2005--2010 median age at marriage
- `marriage`: 2009 marriage rate (per 1000 adults)
- `marriage_se`: Standard error of marriage rate
- `waffle_houses`: Number of Waffle House diners
- `south`: Dummy-coded variable to indicate whether the state is a southern state (0 = non-southern state; 1 = southern state)
- `slaves_1860`: Number of slaves in 1860 census
- `population_1860`: Population from 1860 census
- `prop_slaves_1860`: Proportion of total population that were slaves in 1860



### Preview

```{r}
options(dplyr.width = Inf, width = 85)
waffle = readr::read_csv(file = "~/Documents/github/epsy-8251/data/waffle-house.csv")
head(waffle)
```


### References

The data come from Richard McElreath's **rethinking** package.

McElreath, R. (2016). *rethinking: Statistical Rethinking book package (v. 1.59).*


