---
title: "work-demands.csv"
output: 
  rmdformats::html_docco:
    css: 'codebooks.css'
---


```{r knitr_init, echo=FALSE, cache=FALSE}
library(knitr)
library(rmdformats)
library(fontawesome)

## Global options
options(max.print="75")
opts_chunk$set(echo=FALSE,
	             cache=FALSE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=75)
```


Research has documented the importance of a healthy work-life balance. The data in [work-demands.csv](https://github.com/zief0002/epsy-8251/raw/master/data/work-demands.csv) were simulated to mimic the effects of boundary-spanning work (receiving work-related contact outside of normal work hours) found in the literature on psychological outcomes. The variables in the data are:

- `guilt`: Standardized measure of guilt (as a transitory affective state) related to home-life balance. This was measured using survey questions such as,  "In the past seven days, on how many  days  have  you  felt  guilty?" Higher values indicate more perceived guilt.
- `bound_span_work`: Standardized measure of boundary-spanning work demands. Measured using survey items such as, "how  often  do  coworkers,  supervisors, managers, customers, or clients contact you about work-related  matters  outside  normal  work  hours?". Higher values indicate higher degrees of boundary-spanning work demands.
- `female`: Dummy-coded indicator of sex (0 = Not female; 1 = Female)
- `authority`: Standardized measure of job authority. Measured using survey items such as: "Do you influence or set the rate of pay  received  by  others?" and "Do you have  the authority to hire or fire others?". Higher values indicate more authority.
- `married`: Dummy-coded indicator of marital status (0 = Not married; 1 = Married)



### Preview

```{r}
work = readr::read_csv(file = "~/Documents/github/epsy-8251/data/work-demands.csv")
head(work)
```


### References


Glavin, P., Schieman, S., &amp; Reid, S. (2011). Boundary-spanning work demands and their consequences for guilt and psychological distress. *Journal of Health and Social Behavior, 52*(1) 43&ndash;57. doi: 10.1177/0022146510395023

