---
title: "Some Other Useful(?) R Syntax"
subtitle: "EPsy 8251"
author: Andrew Zieffler"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    lib_dir: "css"
    css: ["default", "metropolis", "metropolis-fonts", "css/extra.css"]
    nature:
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE, servr.daemon = TRUE)
```



# Creating a Data Frame

We create a data frame by using the `data.frame()` function. The arguments for this function are named vectors. Remember to assign your data frame to an object.

```{r}
# Create data frame
clients = data.frame(
  age = c(44, 40, 12, 6),
  sex = c("male", "female", "female", "femle"),
  siblings = c(2, 0, 3, 2)
)

# View data frame
clients
```


# Adding New Variables to a Data Frame

Sometimes you might want to add a new column to an existing data frame. One way to do this is to use the `$` operator in the same way as we used to access an existing column. The only difference is we use this to assign the new column we want to create. 

For example, to create a variable `adult` that indicates whether each client is an adult $(\mathtt{age} \geq 18)$ we use the following syntax:

```{r}
# Create new adult variable
clients$adult = clients$age >= 18

# View data set
clients
```

---


