---
title: "Creating Tables to Present Statistical Results"
date: "`r Sys.Date()`"
output:
  rmdformats::readthedown:
    highlight: kate
    css: table-styles.css
---


```{r knitr_init, echo=FALSE, cache=FALSE}
library(knitr)
library(rmdformats)

## Global options
options(max.print="75")
opts_chunk$set(echo=FALSE,
	             cache=TRUE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=75)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Load libraries
library(corrr)
library(dplyr)
library(knitr)
library(kableExtra)
library(readr)
library(tidyr)

# Read in data
ed = read_csv(file = "~/Documents/github/epsy-8252/data/ed-schools-2018.csv") %>% drop_na()
riverview = read_csv(file = "~/Documents/github/epsy-8252/data/riverview.csv")
```

# Presenting Summary Statistics

Studies often present summary statistics such as the mean and standard deviation. If you only have one or two variables for which you are presenting summary statistics it is often better to present these in the prose of your manuscript. If you have several variables, the information is typically better conveyed in a table. Here is an example of a table presenting summary statistics for a sample of graduate schools of education.

```{r echo=FALSE}
# Set up data frame
data.frame(
  Measure = c("Peer rating", "Acceptance rate for Ph.D. students", "Enrollment", "GRE score (verbal)", "GRE score (quantitative)"),
  M  = c(mean(ed$peer), mean(ed$doc_accept), mean(ed$enroll), mean(ed$gre_verbal), mean(ed$gre_quant)),
  SD = c(sd(ed$peer), sd(ed$doc_accept), sd(ed$enroll), sd(ed$gre_verbal), sd(ed$gre_quant))
) %>%
  kable(
    col.names = c("Measure", "*M*", "*SD*"),
    align = c("l", "r", "r"),
    digits = 1,
    caption = "Table 1<br /><br />*Means and Standard Deviations of Five Measures of Graduate Programs of Education* ($N=122$)"
    ) %>%
  row_spec(row = 0, align = "center")
```

Here are a few things to note about the table:

- It is numbered/named as "Table X". As per the *APA Publication Manual*, this is done in normal text.
- It has a caption. As per the *APA Publication Manual*, this is done in italic text and it uses title-case.
- The different variables being summarized are presented in the table rows. These are given names that readers can understand. (They are **not** the variable names used in the dataset which have shortened names like `doc_accept` and `peer`.)
- The different statistics being used to summarize the data are presented in the table columns. These are given abbreiviated names as suggested in the *APA Publication Manual*. In general, statistics are italicized (this is also done in the text of your manuscript).
- In each column where numbers are presented, the decimal point should be vertically aligned. (Note that not all the variables need to have the same number of decimal places.)
- There are no vertical borders in an APA-formatted table. There are horizontal borders above and below the header row, and at the bottom of the table. Other horizontal borders can be included to help readers if the table is particularly long.

Here is another table presenting the same summarizations, but this time the statistical sumamries are conditioned on sex.


```{r echo=FALSE, eval=FALSE}
# Set up data frame
data.frame(
  Measure = c("Education level (in years)", "Seniority (in years)", "Income (in U.S. dollars"),
  M_f  = c(
    mean(riverview$education[riverview$gender == "female"]),
    mean(riverview$seniority[riverview$gender == "female"]),
    mean(riverview$income[riverview$gender == "female"])
  ),
  SD_f = c(
    sd(riverview$education[riverview$gender == "female"]),
    sd(riverview$seniority[riverview$gender == "female"]),
    sd(riverview$income[riverview$gender == "female"])
    ),
  M_m  = c(
    mean(riverview$education[riverview$gender == "male"]), 
    mean(riverview$seniority[riverview$gender == "male"]),
    mean(riverview$income[riverview$gender == "male"])
    ),
  SD_m = c(
    sd(riverview$education[riverview$gender == "male"]), 
    sd(riverview$seniority[riverview$gender == "male"]),
    sd(riverview$income[riverview$gender == "male"])
    )
) %>%
  kable(
    col.names = c("Measure", "*M*", "*SD*", "*M*", "*SD*"),
    align = c("l", "r", "r", "r", "r"),
    digits = 0,
    caption = "Table 2<br /><br />*Means and Standard Deviations of Three Measures of Riverview Employees Conditioned on Sex*"
    ) %>%
  row_spec(row = 0, align = "center")  %>%
  add_header_above(
    header = c(" " = 1, "Females" = 2, "Males" = 2)
    ) %>%
  save_kable(file = "~/Desktop/creating-tables/table_2.html", self_contained = FALSE)
```

<table>
<caption>Table 2<br><br>*Means and Standard Deviations of Three Measures of Riverview Employees Conditioned on Sex*</caption>
 <thead>
<tr>
<th style="border-bottom:hidden" colspan="1"></th>
<th style="border-bottom:hidden; padding-bottom:0; padding-left:3px;padding-right:3px;text-align: center; " colspan="2"><div style="border-bottom: 1px solid #000; padding-bottom: 5px; ">Females</div></th>
<th style="border-bottom:hidden; padding-bottom:0; padding-left:3px;padding-right:3px;text-align: center; " colspan="2"><div style="border-bottom: 1px solid #000; padding-bottom: 5px; ">Males</div></th>
</tr>
  <tr>
   <th style="text-align:left;text-align: center;"> Measure </th>
   <th style="text-align:right;text-align: center;"> *M* </th>
   <th style="text-align:right;text-align: center;"> *SD* </th>
   <th style="text-align:right;text-align: center;"> *M* </th>
   <th style="text-align:right;text-align: center;"> *SD* </th>
  </tr>
 </thead>
<tbody>
  <tr>
   <td style="text-align:left;"> Education level (in years) </td>
   <td style="text-align:right;"> 16 </td>
   <td style="text-align:right;"> 4 </td>
   <td style="text-align:right;"> 16 </td>
   <td style="text-align:right;"> 5 </td>
  </tr>
  <tr>
   <td style="text-align:left;"> Seniority (in years) </td>
   <td style="text-align:right;"> 14 </td>
   <td style="text-align:right;"> 7 </td>
   <td style="text-align:right;"> 16 </td>
   <td style="text-align:right;"> 7 </td>
  </tr>
  <tr>
   <td style="text-align:left;"> Income (in U.S. dollars </td>
   <td style="text-align:right;"> 48938 </td>
   <td style="text-align:right;"> 13265 </td>
   <td style="text-align:right;"> 59919 </td>
   <td style="text-align:right;"> 14210 </td>
  </tr>
</tbody>
</table>


And one last table.


```{r echo=FALSE, eval=FALSE}
# Set up data frame
data.frame(
  Measure = c("Education level (in years)", "Seniority (in years)", "Income (in U.S. dollars"),
  M_f  = c("16 (4)", "14 (7)", "48938	(13265)"),
  CI_f = c("[13.7, 17.8]", "[10.7, 17.5]", "[42342, 55535]"),
  M_m  = c("16 (5)", "16 (7)", "59919	(14210)"),
  CI_m = c("[13.5, 19.1]", "[11.5, 19.9]", "[51714, 68123]")
) %>%
  kable(
    col.names = c("Measure", "*M* (*SD*)", "95% CI", "*M* (*SD*)", "95% CI"),
    align = c("l", "c", "c", "c", "c"),
    digits = 0,
    caption = "Table 3<br /><br />*Means, Standard Deviations, and Confidence Intervals (CIs) of Three Measures of Riverview Employees Conditioned on Sex*"
    ) %>%
  row_spec(row = 0, align = "center")  %>%
  add_header_above(
    header = c(" " = 1, "Females" = 2, "Males" = 2)
  ) %>%
  save_kable(file = "~/Desktop/creating-tables/table_3.html", self_contained = FALSE)
```

<table>
<caption>Table 3<br><br>*Means, Standard Deviations, and Confidence Intervals (CIs) of Three Measures of Riverview Employees Conditioned on Sex*</caption>
 <thead>
<tr>
<th style="border-bottom:hidden" colspan="1"></th>
<th style="border-bottom:hidden; padding-bottom:0; padding-left:3px;padding-right:3px;text-align: center; " colspan="2"><div style="border-bottom: 1px solid #000; padding-bottom: 5px; ">Females</div></th>
<th style="border-bottom:hidden; padding-bottom:0; padding-left:3px;padding-right:3px;text-align: center; " colspan="2"><div style="border-bottom: 1px solid #000; padding-bottom: 5px; ">Males</div></th>
</tr>
  <tr>
   <th style="text-align:left;text-align: center;"> Measure </th>
   <th style="text-align:center;text-align: center;"> *M* (*SD*) </th>
   <th style="text-align:center;text-align: center;"> 95% CI </th>
   <th style="text-align:center;text-align: center;"> *M* (*SD*) </th>
   <th style="text-align:center;text-align: center;"> 95% CI </th>
  </tr>
 </thead>
<tbody>
  <tr>
   <td style="text-align:left;"> Education level (in years) </td>
   <td style="text-align:center;"> 16 (4) </td>
   <td style="text-align:center;"> [13.7, 17.8] </td>
   <td style="text-align:center;"> 16 (5) </td>
   <td style="text-align:center;"> [13.5, 19.1] </td>
  </tr>
  <tr>
   <td style="text-align:left;"> Seniority (in years) </td>
   <td style="text-align:center;"> 14 (7) </td>
   <td style="text-align:center;"> [10.7, 17.5] </td>
   <td style="text-align:center;"> 16 (7) </td>
   <td style="text-align:center;"> [11.5, 19.9] </td>
  </tr>
  <tr>
   <td style="text-align:left;"> Income (in U.S. dollars </td>
   <td style="text-align:center;"> 48938	(13265) </td>
   <td style="text-align:center;"> [42342, 55535] </td>
   <td style="text-align:center;"> 59919	(14210) </td>
   <td style="text-align:center;"> [51714, 68123] </td>
  </tr>
</tbody>
</table>


# Presenting Correlation Coefficients

Similar to presenting summary statistics, if you only have one or two correlation coefficients to present, it is best to present these in the prose of your manuscript. If you have several correlations the information is typically better conveyed in a table. Here is an example of a table presenting correlation coefficients for our sample of graduate schools of education.

```{r echo=FALSE}
ed %>%
  select(peer, doc_accept, enroll, gre_verbal, gre_quant) %>%
  correlate() %>%
  shave(upper = TRUE) %>%
  fashion(decimals = 2, na_print = "—") %>%
  mutate(
    rowname = c("1. Peer rating", "2. Acceptance rate for Ph.D. students", "3. Enrollment", "4. GRE score (verbal)", "5. GRE score (quantitative)")
  ) %>%
  kable(
    caption = "Table 4<br /><br />*Intercorrelations between Five Measures of Graduate Programs of Education*",
    col.names = c("Measure", "1", "2", "3", "4", "5"),
    align = c("l", "r", "r", "r", "r", "r")
  ) %>%
  row_spec(row = 0, align = "center")
```

Here are a few things to note about the table:

- The correlation coefficients are generally rounded to two decimal places.
- In each column where numbers are presented, the decimal point should be vertically aligned.
- If the correlation table is to support a regression analysis, typically the outcome variable is the first variable presented in the table (in this case, peer rating). If there is a focal predictor (i.e., a predictor germaine to your research question), this should be the second variable presented in the table, etc. Otherwise present the predictors alphabetically.
- p-values should not be presented in a table of correlations. There are many issues related to p-values in a table of correlations, not the least of which is that of multiple tests. It is better to save any presentation of p-values for tables of the regression results.
- Do not indicate "statistical significance" with stars as per the recommendation of the *American Statistical Association*.

Here is an alternative table presenting both the summary statistics of each variable and the intercorrelations. Combining the information into a single table can be useful when trying to save space in a manuscript.

```{r echo=FALSE}
tab_05 = ed %>%
  select(peer, doc_accept, enroll, gre_verbal, gre_quant) %>%
  correlate() %>%
  shave(upper = TRUE) %>%
  fashion(decimals = 2, na_print = "—")

tab_05[1, 2] = "3.3 (0.5)"
tab_05[2, 3] = "40.1 (20.2)" 
tab_05[3, 4] = "970 (665)" 
tab_05[4, 5] = "154.9 (3.7)" 
tab_05[5, 6] = "151.0 (4.4)" 
  
tab_05 %>% 
  mutate(
    rowname = c("1. Peer rating", "2. Acceptance rate for Ph.D. students", "3. Enrollment", "4. GRE score (verbal)", "5. GRE score (quantitative)")
  ) %>%
  kable(
    caption = "Table 5<br /><br />*Intercorrelations between Five Measures of Graduate Programs of Education. Means and Standard Deviations are Presented Along the Main Diagonal*",
    col.names = c("Measure", "1", "2", "3", "4", "5"),
    align = c("l", "c", "c", "c", "c", "c")
  ) %>%
  row_spec(row = 0, align = "center")
```

# Presenting Results from a Fitted Regression Model

Typically the results of the "final" adopted model are presented in a table. However, if there are only one or two predictors in the model, it is best to present these in the prose of your manuscript rather than a table. If you have several predictors the information is often better conveyed in a table. Here is an example of a table presenting the results from a fitted regression model for our sample of graduate schools of education.

```{r echo=FALSE}
tab_06 = lm(peer ~ 1 + doc_accept + enroll + gre_verbal + gre_quant, data = ed) %>%
  broom::tidy() %>% 
  mutate(
    term = c("Constant", "Acceptance rate for Ph.D. students", "Enrollment", "GRE score (verbal)", "GRE score (quantitative)")
  ) 

tab_06 = tab_06[c(2:5, 1), ]

tab_06 %>%
  kable(
    caption = "Table 6<br /><br />*Coefficient-Level Estimates for a Model Fitted to Estimate Variation in Peer Ratings ($N=129$; $R^2=0.387$)*",
    col.names = c("Predictor", "*B*", "*SE*", "*t*", "*p*"),
    digits = c(0, 2, 3, 2, 3),
    align = c("l", "r", "r", "r", "r")
  ) %>%
  row_spec(row = 0, align = "center")
```

Here are a few things to note about the table:

- The intercept term of the model (Constant), is relegated to the bootom of the table as it is not typically susbstantively interesting.
- The different coefficients are presented in the table rows (when the table includes only one model). These are given names that readers can understand. (They are **not** the variable names used in the dataset which have shortened names like `doc_accept` and `peer`.)
- There are no stars indicating statistical significance as per the *American Statistical Association's* recommendation.
- The sample size ($N$) and variance accounted for ($R^2$) estimates could be added to a footnote instead of the caption.


Alternatively, a regression table can include the confidence interval for each of the coefficients in addition to (or in place of!) the *p*-values. This addresses the uncertainty in the estimates.

```{r echo=FALSE}
tab_06 %>%
  mutate(
    CI = c("[-0.014, -0.006]", "[0.000, 0.000]", "[-0.031, 0.034]", "[0.007, 0.061]", "[-4.894, 1.521]"),
    Beta = c(-.43, .07, .01, .32, .01)
  ) %>%
  select(term, estimate, std.error, CI)  %>%
  kable(
    caption = "Table 7<br /><br />*Coefficient-Level Estimates and Uncertainty for a Model Fitted to Estimate Variation in Peer Ratings ($N=129$; $R^2=0.387$)*",
    col.names = c("Predictor", "*B*", "*SE*", "95% CI"),
    digits = c(0, 2, 2, 0),
    align = c("l", "r", "r", "r")
  ) %>%
  row_spec(row = 0, align = "center")
```


Another variation on this table includes the standardized regression coefficients.



```{r echo=FALSE}
tab_06 %>%
  mutate(
    CI = c("[-0.014, -0.006]", "[0.000, 0.000]", "[-0.031, 0.034]", "[0.007, 0.061]", "[-4.894, 1.521]"),
    Beta = c(-.43, .07, .01, .32, .01)
  ) %>%
  select(term, estimate, std.error, CI, Beta)  %>%
  kable(
    caption = "Table 8<br /><br />*Coefficient-Level Estimates (Unstandardized and Standardized) and Uncertainty for a Model Fitted to Estimate Variation in Peer Ratings ($N=129$; $R^2=0.387$)*",
    col.names = c("Predictor", "*B*", "*SE*", "95% CI", "$\\beta$"),
    digits = c(0, 2, 2, 0, 2),
    align = c("l", "r", "r", "r", "r")
  ) %>%
  row_spec(row = 0, align = "center")
```


# Presenting Results from Many Fitted Regression Models

In many analyses, you may need to present the results from a set of fitted models. Here is an example of a table presenting the results from a set of fitted regression models for our sample of graduate schools of education.


```{r, message=FALSE, echo=FALSE, results='asis', eval=FALSE}
library(stargazer)

lm.1 = lm(peer ~ 1 + gre_verbal + gre_quant, data = ed)
lm.2 = lm(peer ~ 1 + doc_accept + enroll, data = ed)
lm.3 = lm(peer ~ 1 + doc_accept + enroll + gre_verbal + gre_quant, data = ed)

stargazer(lm.1, lm.2, lm.3,
  ci = TRUE,
  column.labels = c("Model 1", "Model 2", "Model 3"),
  covariate.labels = c("GRE score (verbal)", "GRE score (quantitative)", "Acceptance rate for Ph.D. students", "Enrollment"),
  dep.var.caption = "Outcome variable: Peer ratings",
  dep.var.labels = NULL,
  dep.var.labels.include = FALSE,
  type = "html",
  keep.stat = c("rsq","ser"),
  star.cutoffs = NA,
  header = FALSE,
  table.placement = 'H',
  title = "Unstandardized Coefficients and Confidence Intervals for a Series of Regression Models Fitted to Data from $n=129$ Graduate Schools of Education to Predict Variation in Peer Ratings",
  omit.table.layout = "n"
  )
```


<table style="text-align:center">
<caption>Table 9<br /><br /><i>Unstandardized Coefficients and Confidence Intervals from a Series of Regression Models Fitted to Predict Variation in Peer Ratings for n=129 Graduate Schools of Education</i></caption>
<tr><td colspan="4" style="border-bottom: 1px solid black"></td></tr>
<tr><td style="text-align:left; border-bottom: 1px solid black"></td><td style="border-bottom: 1px solid black">Model 1</td><td style="border-bottom: 1px solid black">Model 2</td><td style="border-bottom: 1px solid black">Model 3</td></tr>
<tr><td style="text-align:left">GRE score (verbal)</td><td>0.011</td><td></td><td>0.001</td></tr>
<tr><td style="text-align:left"></td><td>(-0.024, 0.046)</td><td></td><td>(-0.031, 0.033)</td></tr>
<tr><td style="text-align:left"></td><td></td><td></td><td></td></tr>
<tr><td style="text-align:left">GRE score (quantitative)</td><td>0.047</td><td></td><td>0.036</td></tr>
<tr><td style="text-align:left"></td><td>(0.017, 0.076)</td><td></td><td>(0.009, 0.063)</td></tr>
<tr><td style="text-align:left"></td><td></td><td></td><td></td></tr>
<tr><td style="text-align:left">Acceptance rate for Ph.D. students</td><td></td><td>-0.013</td><td>-0.010</td></tr>
<tr><td style="text-align:left"></td><td></td><td>(-0.017, -0.009)</td><td>(-0.014, -0.006)</td></tr>
<tr><td style="text-align:left"></td><td></td><td></td><td></td></tr>
<tr><td style="text-align:left">Enrollment</td><td></td><td>0.0001</td><td>0.0001</td></tr>
<tr><td style="text-align:left"></td><td></td><td>(-0.00004, 0.0002)</td><td>(-0.0001, 0.0002)</td></tr>
<tr><td style="text-align:left"></td><td></td><td></td><td></td></tr>
<tr><td style="text-align:left">Constant</td><td>-5.488</td><td>3.769</td><td>-1.857</td></tr>
<tr><td style="text-align:left"></td><td>(-8.683, -2.294)</td><td>(3.572, 3.967)</td><td>(-5.054, 1.340)</td></tr>
<tr><td style="text-align:left"></td><td></td><td></td><td></td></tr>
<tr><td colspan="4" style="border-bottom: 1px solid black"></td></tr><tr><td style="text-align:left">R<sup>2</sup></td><td>0.243</td><td>0.300</td><td>0.390</td></tr>
<tr><td style="text-align:left">Residual Std. Error</td><td>0.429</td><td>0.413</td><td>0.389</td></tr>
<tr><td colspan="4" style="border-bottom: 1px solid black"></td></tr></table>


Here are a few things to note about the table:

- When presenting the results from multiple models, the goal is to often compare how coefficients differ from model-to-model. Because of this we typically present each model in a column and each coefficient in a row&emdash;this way you compare by reading horizontally.
- If a model does not include a particular coefficient, leave that cell blank.
- The intercept term of the model (Constant), is relegated to the bootom of the table as it is not typically susbstantively interesting.
- Use predictor names that readers can understand. (They are **not** the variable names used in the dataset which have shortened names like `doc_accept` and `peer`.)
- Model-level estimates are also included in the table, typically below the coefficient-level estimates. Sometimes a horizontal line is added to the table as a separator.
- The sample size ($N$) could be added to a footnote instead of the caption.

If you have many models to present, use a landscape orientation on your page.


If you must present *p*-values, do not include stars indicating statistical significance (as per the *American Statistical Association's* recommendation).

```{r, message=FALSE, echo=FALSE, results='asis', eval=FALSE}
stargazer(lm.1, lm.2, lm.3,
  column.labels = c("Model 1", "Model 2", "Model 3"),
  covariate.labels = c("GRE score (verbal)", "GRE score (quantitative)", "Acceptance rate for Ph.D. students", "Enrollment"),
  dep.var.caption = "Outcome variable: Peer ratings",
  dep.var.labels = NULL,
  dep.var.labels.include = FALSE,
  type = "html",
  keep.stat = c("rsq","ser"),
  report = c("vcsp"),
  star.cutoffs = NA,
  header = FALSE,
  table.placement = 'H',
  title = "Unstandardized Coefficients and Confidence Intervals for a Series of Regression Models Fitted to Data from $n=129$ Graduate Schools of Education to Predict Variation in Peer Ratings",
  omit.table.layout = "n"
  )
```


<table style="text-align:center">
<caption>Table 10<br /><br /><i>Unstandardized Coefficients (Standard Errors) and p-Values for a Series of Regression Models Fitted to Data from n=129 Graduate Schools of Education to Predict Variation in Peer Ratings</i></caption>
<tr><td colspan="4" style="border-bottom: 1px solid black"></td></tr>
<tr><td style="text-align:left; border-bottom: 1px solid black"></td><td style="border-bottom: 1px solid black">Model 1</td><td style="border-bottom: 1px solid black">Model 2</td><td style="border-bottom: 1px solid black">Model 3</td></tr>
<tr><td style="text-align:left">GRE score (verbal)</td><td>0.011</td><td></td><td>0.001</td></tr>
<tr><td style="text-align:left"></td><td>(0.018)</td><td></td><td>(0.016)</td></tr>
<tr><td style="text-align:left"></td><td>p = 0.531</td><td></td><td>p = 0.950</td></tr>
<tr><td style="text-align:left"></td><td></td><td></td><td></td></tr>
<tr><td style="text-align:left">GRE score (quantitative)</td><td>0.047</td><td></td><td>0.036</td></tr>
<tr><td style="text-align:left"></td><td>(0.015)</td><td></td><td>(0.014)</td></tr>
<tr><td style="text-align:left"></td><td>p = 0.003</td><td></td><td>p = 0.012</td></tr>
<tr><td style="text-align:left"></td><td></td><td></td><td></td></tr>
<tr><td style="text-align:left">Acceptance rate for Ph.D. students</td><td></td><td>-0.013</td><td>-0.010</td></tr>
<tr><td style="text-align:left"></td><td></td><td>(0.002)</td><td>(0.002)</td></tr>
<tr><td style="text-align:left"></td><td></td><td>p = 0.000</td><td>p = 0.00000</td></tr>
<tr><td style="text-align:left"></td><td></td><td></td><td></td></tr>
<tr><td style="text-align:left">Enrollment</td><td></td><td>0.0001</td><td>0.0001</td></tr>
<tr><td style="text-align:left"></td><td></td><td>(0.0001)</td><td>(0.0001)</td></tr>
<tr><td style="text-align:left"></td><td></td><td>p = 0.239</td><td>p = 0.347</td></tr>
<tr><td style="text-align:left"></td><td></td><td></td><td></td></tr>
<tr><td style="text-align:left">Constant</td><td>-5.488</td><td>3.769</td><td>-1.857</td></tr>
<tr><td style="text-align:left"></td><td>(1.630)</td><td>(0.101)</td><td>(1.631)</td></tr>
<tr><td style="text-align:left"></td><td>p = 0.002</td><td>p = 0.000</td><td>p = 0.258</td></tr>
<tr><td style="text-align:left"></td><td></td><td></td><td></td></tr>
<tr><td colspan="4" style="border-bottom: 1px solid black"></td></tr><tr><td style="text-align:left">R<sup>2</sup></td><td>0.243</td><td>0.300</td><td>0.390</td></tr>
<tr><td style="text-align:left">Residual Std. Error</td><td>0.429</td><td>0.413</td><td>0.389</td></tr>
<tr><td colspan="4" style="border-bottom: 1px solid black"></td></tr></table>


An alternative presentation puts the coefficient and standard error in the same row.

<table style="text-align:center">
<caption>Table 11<br /><br /><i>Unstandardized Coefficients (Standard Errors) and p-Values for a Series of Regression Models Fitted to Data from n=129 Graduate Schools of Education to Predict Variation in Peer Ratings</i></caption>
<tr><td colspan="4" style="border-bottom: 1px solid black"></td></tr>
<tr><td style="text-align:left; border-bottom: 1px solid black"></td><td style="border-bottom: 1px solid black">Model 1</td><td style="border-bottom: 1px solid black">Model 2</td><td style="border-bottom: 1px solid black">Model 3</td></tr>
<tr><td style="text-align:left">GRE score (verbal)</td><td>0.011 (0.018)</td><td></td><td>0.001 (0.016)</td></tr>
<tr><td style="text-align:left"></td><td>p = 0.531</td><td></td><td>p = 0.950</td></tr>
<tr><td style="text-align:left"></td><td></td><td></td><td></td></tr>

<tr><td style="text-align:left">GRE score (quantitative)</td><td>0.047 (0.015)</td><td></td><td>0.036 (0.014)</td></tr>
<tr><td style="text-align:left"></td><td>p = 0.003</td><td></td><td>p = 0.012</td></tr>
<tr><td style="text-align:left"></td><td></td><td></td><td></td></tr>

<tr><td style="text-align:left">Acceptance rate for Ph.D. students</td><td></td><td>-0.013 (0.002)</td><td>-0.010 (0.002)</td></tr>
<tr><td style="text-align:left"></td><td></td><td>p = 0.000</td><td>p = 0.00000</td></tr>
<tr><td style="text-align:left"></td><td></td><td></td><td></td></tr>

<tr><td style="text-align:left">Enrollment</td><td></td><td>0.0001 (0.0001)</td><td>0.0001 (0.0001)</td></tr>
<tr><td style="text-align:left"></td><td></td><td>p = 0.239</td><td>p = 0.347</td></tr>
<tr><td style="text-align:left"></td><td></td><td></td><td></td></tr>

<tr><td style="text-align:left">Constant</td><td>-5.488 (1.630)</td><td>3.769 (0.101)</td><td>-1.857 (1.631)</td></tr>
<tr><td style="text-align:left"></td><td>p = 0.002</td><td>p = 0.000</td><td>p = 0.258</td></tr>
<tr><td style="text-align:left"></td><td></td><td></td><td></td></tr>

<tr><td colspan="4" style="border-bottom: 1px solid black"></td></tr><tr><td style="text-align:left">R<sup>2</sup></td><td>0.243</td><td>0.300</td><td>0.390</td></tr>
<tr><td style="text-align:left">Residual Std. Error</td><td>0.429</td><td>0.413</td><td>0.389</td></tr>
<tr><td colspan="4" style="border-bottom: 1px solid black"></td></tr></table>



