---
title: "EPsy 8251"
output:
  rmdformats::material:
    highlight: kate
    css: website-style.css
    includes:
      in_header: header.html
---


```{r knitr_init, echo=FALSE, cache=FALSE}
library(knitr)
library(rmdformats)
library(fontawesome)

## Global options
options(max.print="75")
opts_chunk$set(echo=FALSE,
	             cache=TRUE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=75)
```


# Welcome


Welcome to *EPsy 8251: Methods in Data Analysis for Educational Research I*. This is the first course of a two-semester statistical methods sequence for doctoral students in education.  

<!-- - **Syllabus, assignments, data, and script files:** [github repository](https://github.com/zief0002/epsy-8251) -->
<!-- - **Course notes and readings:** [website](https://zief0002.github.io/book-8252/) | [github repository](https://github.com/zief0002/book-8252) -->


<br />

### Instructor {-}

Andrew Zieffler ([zief0002@umn.edu](mailto://zief0002@umn.edu)) <br />
**Office:** [Education Sciences Building 178](https://www.google.com/maps/place/Education+Sciences+Building/@44.9784043,-93.2394586,15z/data=!4m2!3m1!1s0x0:0x45656dac481b9150)  <br />
**Office Hours:** Tuesday 9:00 AM&ndash;10:00 AM; and by appointment


### Teaching Assistant {-}

Mireya Smith ([mart1799@umn.edu](mailto://mart1799@umn.edu))  <br />
**Office:** [Education Sciences Building 192](https://www.google.com/maps/place/Education+Sciences+Building/@44.9784043,-93.2394586,15z/data=!4m2!3m1!1s0x0:0x45656dac481b9150)  <br />
**Office Hours:** Thursdays 9:00 AM&ndash;11:00 AM; and by appointment

<br />

### Classroom

- Monday/Wednesday (2:30&ndash;3:45): [Ford Hall 150](http://campusmaps.umn.edu/ford-hall)
- Tuesday/Thursday (11:15&ndash;12:30): [Ford Hall B15](http://campusmaps.umn.edu/ford-hall)

<br />


### Course Content and Syllabus {-}

The statistical content for EPsy 8251 includes rigorous coverage of estimation and hypothesis testing with a particular focus on ANOVA and multiple linear regression.

- The course syllabus is available  [here](https://github.com/zief0002/epsy-8251/raw/master/syllabus/f19-8251-syllabus.pdf).

<br />

### Textbooks {-}

The primary course textbook is available via the University of Minnesota library.

- [Applied regression: An introduction](https://primo.lib.umn.edu/primo-explore/fulldisplay?docid=UMN_ALMA21533776460001701&context=L&vid=TWINCITIES&search_scope=mncat_discovery&tab=article_discovery&lang=en_US) [required]

The secondary course textbook (available via Amazon) is

- [Presenting your findings: A practical guide for creating tables](https://www.amazon.com/Presenting-Your-Findings-Practical-Creating/dp/143380705X) [required] 

<br />


```{r child = 'schedule.Rmd'}
```


```{r child = 'assignments.Rmd'}
```


```{r child = 'data.Rmd'}
```







